---
title: "Getting Started with etlspatial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with etlspatial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(etlspatial)
library(sf)
library(DBI)
library(duckdb)
```

`etlspatial` is a lightweight spatial ETL framework for building reproducible, production-ready geospatial workflows in R.

It provides a consistent interface for reading, standardising, and writing spatial data across ESRI formats (file geodatabases and shapefiles) and open formats such as GeoPackage, with integrated DuckDB support for scalable analytics.

The package is designed to be engine-agnostic, portable, and robust to changes in data source environments.

## Spatial Format Support

`etlspatial` supports both ESRI-native and open spatial formats:

- ESRI File Geodatabases (`.gdb`)
- ESRI Shapefiles (`.shp`)
- GeoPackage (`.gpkg`)

GeoPackage is fully supported as a portable, open standard, enabling workflows to run in environments where ESRI-specific formats are not available.

All formats are handled through a consistent interface, allowing ETL workflows to remain stable regardless of the underlying storage type.

## Load Demo Data

A small, package-safe demo dataset is included to demonstrate core functionality.

The dataset contains ABS SA4 boundaries for Victoria (GDA2020), stored as a GeoPackage.

```{r Load Demo}

# Load demo data

demo_path <- system.file(
  "extdata",
  "abs_sa4_vic_demo.gpkg",
  package = "etlspatial"
)

sa4 <- read_esri_layer(
  dsn = demo_path,
  layer = "abs_sa4_vic_demo",
  quiet = TRUE
)

dplyr::glimpse(sa4, width = 80)
```

## QA Check

Run a quick spatial summary:

```{r qa-check}
qa_spatial_summary(sa4)
```

Optional quick plot:

```{r quick-plot, fig.width=7, fig.height=5}
qa_spatial_plot(sa4)
```

## Create DuckDB Connection

The dataset is written to DuckDB using WKT-based geometry storage.

Create a temporary DuckDB database:

```{r create-duckdb}
duckdb_path <- tempfile(fileext = ".duckdb")

con <- DBI::dbConnect(
  duckdb::duckdb(),
  dbdir = duckdb_path
)
```

## Write to DuckDB

```{r write-to-duckdb}
write_sf_to_duckdb(
  x = sa4,
  con = con,
  table_name = "sa4_vic"
)
```

## Read from DuckDB

Reconstruct the `sf` object:

```{r read-from-duckdb}
sa4_from_db <- read_sf_from_duckdb(
  con = con,
  table_name = "sa4_vic"
)

dplyr::glimpse(sa4_from_db, width = 80)
```

## Check DuckDB Registry

View ETL registry:

```{r check-registry}
etl_duckdb_registry(duckdb_path)
```

## Write Back to Spatial Format

Example: write to GeoPackage.

```{r write-gpkg, results='hide'}

sa4_from_db <- sf::st_set_crs(sa4_from_db, 7844)

out_path <- tempfile(fileext = ".gpkg")

write_esri_layer(
  x = sa4_from_db,
  dsn = out_path,
  layer = "sa4_vic_output",
  quiet = TRUE
)
```

```{r show-output-path}
cat("Output file:", basename(out_path))
```

## Example: Continue the Workflow with `sf`

Once data is written to DuckDB and read back as an `sf` object, it can be used directly in normal spatial workflows.

This example creates centroid points from the SA4 polygons and writes the result back to DuckDB.

```{r centroid-example, warning=FALSE}
# Ensure CRS is correctly assigned after reading from DuckDB
sa4_from_db <- sf::st_set_crs(sa4_from_db, 7844)

# Validate and normalise geometry for this processing step
sa4_clean <- sa4_from_db |>
  sf::st_make_valid()

# Create representative points safely inside each polygon
sa4_points <- sa4_clean |>
  sf::st_point_on_surface()

sf::st_crs(sa4_points)$epsg

dplyr::glimpse(sa4_points, width = 80)
```

Write the centroid layer to DuckDB.

```{r write-points, results='hide'}
write_sf_to_duckdb(
  x = sa4_points,
  con = con,
  table_name = "sa4_vic_points",
  overwrite = TRUE
)
```

Confirm the new layer is tracked in the DuckDB registry.

```{r registry-after-centroids}
etl_duckdb_registry(duckdb_path)
```

## Close DuckDB Connection

```{r close-duckdb}
DBI::dbDisconnect(con, shutdown = TRUE)
```

## Summary

This workflow demonstrates:

- Reading spatial data from GeoPackage
- Standardising and validating geometry
- Writing and retrieving spatial data via DuckDB
- Continuing spatial analysis using `sf`

The package supports reproducible, portable spatial ETL workflows for:

- Spatial analytics
- Risk modelling
- Spatial epidemiology
- Large-scale geospatial processing